Calculating Module Enrichment and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}

### 2016-01-11 jss/article.tex
[skip ci] gbm example update

\section[Package organization]{Packages organization}

@@

@@

\begin{figure}[!ht]
  \caption{Mean expression values for ER negative differentially expressed
  genes in breast cancer visualized as heatmaps on the ACSN master map.} 
  \centering
  \includegraphics[width=0.8\textwidth]{figures/gbm.pdf}
  \label{fig:mainz}
\end{figure}

>>>>

\section[Packages organization]{Packages organization}

@@

Recent years have witnessed a dramatic increase in new technologies for
interrogating the activity levels of various cellular components on a
genome-wide scale, including genomic, epigenomic, transcriptomic, and proteomic
information \citep{hawkins2010next}. Integrating these heterogeneous datasets
provides more biological insights than performing separate analyses. For
instance, international consortia such as The Cancer Genome Atlas (TCGA) have
launched large-scale initiatives to characterize multiple types of cancer at
different levels on hundreds of samples.  These integrative studies have
already led to the identification of novel cancer genes
\citep{mclendon2008comprehensive}.

Malignant gliomas, the most common subtype of primary brain tumors, are
aggressive, highly invasive, and neurologically destructive tumors considered to
be among the deadliest of human cancers. In its most aggressive manifestation,
glioblastoma (GBM), median survival ranges from 9 to 12 months, despite maximum
treatment efforts \citep{maher2001malignant}. In this study we have analyzed
whole-genome mutation data generated by the TCGA project on hundreds of
patients. More specifically, we parsed the MAF (Mutation Annotation Format) GBM
files produced by different sequencing centers to count and calculate gene
mutation frequencies. We kept the mutations having a status likely to disturb
the target protein's function (i.e, Frame\_Shift\_Del, Frame\_Shift\_Ins,
In\_Frame\_Del, In\_Frame\_Ins, Missense\_Mutation, Nonsense\_Mutation). In
total, we collected mutations for xxx genes in a total of xxx samples. In order
to retain the most frequently mutated genes, we calculated frequencies across
all mutated samples, and kept genes having a frequency greater than 1\%. We
further labelled genes having a frequency greater than 1\% and less than 5\% as "1"
and genes highly mutated (frequency higher than 5\%) as "2".

We loaded the data as a matrix in R and calculated the enrichment in ACSN maps
with the ACSNMineR function \code{enrichment}. The results are displayed in
table \ref{tab:table_gbm}. There are 16 modules significantly enriched in the
Cell Cycle, DNA repair, EMT motility and Survival. Cell matrix adhesions and ECM
(extra cellular matrix) are the modules with highest significance (corrected
p-values of 2.55e-11 and 2.34e-9 respectively). Visualization of the list of
frequently mutated genes is shown on figure \ref{fig:gbm}. This figure was
generated with the ACSNMineR commands detailed below.

@@

\begin{verbatim}
library(RNaviCell)

# Create a NaviCell object, point it to the ACSN master map and launch
# a session.

navicell <- NaviCell()
navicell$proxy_url <- "https://acsn.curie.fr/cgi-bin/nv_proxy.php"
navicell$map_url <- "https://acsn.curie.fr/navicell/maps/acsn/master/index.php"
navicell$launchBrowser()

# Read the GBM data file and import it into the session.

mat <- navicell$readDatatable('gbm.txt')
navicell$importDatatable("Mutation data", "GBM", mat)

# set datatable and sample names for the glyph editor

navicell$drawingConfigSelectGlyph(1, TRUE)
navicell$glyphEditorSelectSample(1, "categ")
navicell$glyphEditorSelectShapeDatatable(1, "GBM")
navicell$glyphEditorSelectColorDatatable(1, "GBM")
navicell$glyphEditorSelectSizeDatatable(1, "GBM")
navicell$glyphEditorApply(1)

# set color, shape and size parameters for glyphs

navicell$unorderedConfigSetDiscreteShape("GBM", "sample", 0, 1)
navicell$unorderedConfigSetDiscreteShape("GBM", "sample", 1, 5)
navicell$unorderedConfigApply("GBM", "shape")

navicell$unorderedConfigSetDiscreteColor("GBM", "sample", 0, "398BC3")
navicell$unorderedConfigSetDiscreteColor("GBM", "sample", 1, "CC5746")
navicell$unorderedConfigApply("GBM", "color")

navicell$unorderedConfigSetDiscreteSize("GBM", "sample", 0, 4)
navicell$unorderedConfigSetDiscreteSize("GBM", "sample", 1, 14)

navicell$unorderedConfigApply("GBM", "size")
\end{verbatim}

\begin{figure}[!ht]
  \caption{Mean expression values for ER negative differentially expressed
  genes in breast cancer visualized as heatmaps on the ACSN master map.} 
  \centering
  \includegraphics[width=0.8\textwidth]{figures/gbm.pdf}
  \label{fig:gbm}
\end{figure}
